Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-11 doc/paper-lcr.tex
more halo stuff

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW profiles \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are less concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. $\rho_0$ is calculated 
from the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee
Defining $x$ as the dimensionless projected radial distance $R/r_{s}$ , \citet{bartelmann1996} derives the projected mass density, which is given by:
\be\label{eq:kappanfw}
\Sigma_{\rm NFW}(x) = 
\begin{cases}
  \frac{2r_{s}\delta_{c}\rho_{c}}{x^{2}-1}
  \left(1-\frac{2}{\sqrt{1-x^{2}}}{\rm arctanh}\sqrt{\frac{1-x}{1+x}}\right)\hspace{0.15cm}
  & \left(x < 1\right) \\
\\
  \frac{2r_{s}\delta_{c}\rho_{c}}{3}
  & \left(x = 1\right)\\
\\
  \frac{2r_{s}\delta_{c}\rho_{c}}{x^{2}-1}
  \left(1-\frac{2}{\sqrt{x^{2}-1}}\arctan\sqrt{\frac{x-1}{1+x}}
  \right) 
  & \left(x > 1\right) 
\end{cases}
\ee
and the convergence is given by $\Sigma_{\rm NFW}(x) / \Sigma_{\rm cr}(z)$.

Disturbingly, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line sight will also be divergant\footnote{At large radius, $\Sigma_{\rm nfw} \propto R^{-2}$, but the differential number of halos centred within an annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$, so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which diverges logarithmically}.
Since infinite mass is unphysical, the profile must be truncated at some point. \flag{previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low; \citet{Nandra2012} showed that the Universe's expansion can place an upper limit on the radius of NFW halos -- for cluster scale halos, this happens at around 25 times the NFW scale radius. In \Sref{sec:knownMh+z} we investigate how different truncations affect the recovered total convergence for a line of sight, using the truncated NFW profile of \citet{BMO}
\be\label{eq:bmoprofile}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee

\flag{kappa truncated NFW}

Not all DM is in halo catalog. Smooth component approximation.  How voids
produce a negative kappa - filled beam distances require halos and voids to
provide density contrast relative to mean. How we correct for voids. Comments on
reasonable, but unphysical nature of this approach, tests to follow.

\flag{kappa smooth correction.}

Thus the total $\kappax$ for a line of sight is given by 
\be
\kappax = \left(\sum_{halos} \kappa_{halos}\right)-\kappa{smooth}.
\ee

@@

For every halo and subhalo in the catalogue, we estimate a concentration according to
the best fit relation of \citet{neto2007}:
\be
c_{200} = 4.67 (M_{200}/10^{14} h^{-1} \Msun)^{-0.11}
\ee

Lines of sight are then chosen from random positions in the catalogue, and for each line of
sight we reconstruct the convergences from every halo within 5 arcminutes and compare the reconstructed
kappas with those from the ray tracing results of Hilbert et al. We do this for a range of
truncations, as a function of both the virial and NFW scale radius. The scatter is 
shown in Figure \ref{fig:ScattervsTruncation}. It is clear from Figure \ref{fig:ScattervsTruncation} 
that as long as the truncation is greater than the virial radius, the choice of truncation
radius does not broadly affect the scatter on $\kappah$-$kapparec$; from now on we choose 
to truncate the halos at three times their virial radius.

>>>>

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW profiles \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are less concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. The best fit relation of \citet{neto2007} is given by
\be
c_{200} = 4.67 (M_{200}/10^{14} h^{-1} \Msun)^{-0.11}.
\ee
$\rho_0$ is calculated from the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee

Disturbingly, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line sight will also be divergant\footnote{At large radius, $\Sigma_{\rm nfw} \propto R^{-2}$, but the differential number of halos centred within an annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$, so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which diverges logarithmically}.Since infinite mass is unphysical, the profile must be truncated at some point. \flag{previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low; \citet{Nandra2012} showed that the Universe's expansion can place an upper limit on the radius of NFW halos -- for cluster scale halos, this happens at around 25 times the NFW scale radius. In \Sref{sec:knownMh+z} we investigate how different truncations affect the recovered total convergence for a line of sight, using the truncated NFW profile of \citet{BMO}
\be\label{eq:bmoprofile}
\rho(r) = 
\frac{\rho_{\rm NFW} (r)}{1+\left(r/r_{t}\right)^2},
\ee
which is the same as the NFW profile in the limit that $r_t$ goes to infinity. Defining $x$ as the dimensionless projected radial distance $R/r_{s}$ and $\tau$ as the dimensionless truncation radius $r_{t}/r_{s}$, \citet{BMO} derives the projected mass density, which is given by:
\begin{align}\label{eq:kappaBMO}
\Sigma_{\rm BMO}(x) = {2\tau^2 \over (\tau^2+1)^2}\left(
        {\tau^2+1\over x^2-1}\left(1-\mathcal{F}(x)\right)
        +
        2\mathcal{F}(x)\right.\nonumber\\
        -
        \left. {\pi \over \sqrt{\tau^2+x^2}}
        +
        {(\tau^2-1)\mathcal{L}(x,\tau)
        \over
        \tau\sqrt{\tau^2+x^2}}\right)
\end{align}
where $\mathcal{F}(x)$ and $\mathcal{L}(x,\tau)$ are defined as
\be\label{eq:F} 
\mathcal{F}(x) \equiv \begin{cases}  \frac{{\rm cos}^{-1} (1/x)}{\sqrt{x^2-1}} \hspace{0.2cm} & (x>1) \\
\\
\frac{4-x^2}{3}  & (x=1)\\
\\
\frac{ {\rm cosh}^{-1} (1/x)}{\sqrt{1-x^2}} & (x<1)
\end{cases}
\ee
\be\label{eq:L}
\mathcal{L}(x,\tau) = \ln\left(\frac{x}{\sqrt{\tau^2+x^2}+\tau}\right)
\ee
%the convergence is given by $\Sigma_{\rm BMO}(x) / \Sigma_{\rm cr}(z)$.

Whilst halos contribute a positive convergence to $\kappax$, voids contribute a negative convergence. When a raybundle passes through a 
region of space that is less dense than $\rho_{\rm cr}(z)$ the rays are de-focused\footnote{In principle, the convergence caused by halos, is also caused by the overdensity, but we neglect this effect since halos are typically much more dense than the intergalactic medium}. Whilst the 
full raytraced solution takes into account the effect of voids, our reconstruction cannot -- the halo catalogue does not include voids. In principle the absence of a halo could be taken to infer a void, but the under-density of such a void is hard to infer. Instead, we shift our $\kappax$ values so that $\langle\kappax\rangle=0$. We refer to this corrective term as $\kappav$. The void correction introduces a source of error, since the inferred value of $\kappav$ will depend on the properties of the sightlines reconstructed, but this error shrinks with the number of sightlines reconstructed. To avoid potential biases we calculate $\kappav$ from randomly selected sightlines. The total $\kappax$ along the line of sight is given by 
\be \label{eq:totalkappax}
\kappax = \left(\sum_{\rm halos} \kappa_{\rm halo}\right)-\kappav.
\ee

@@

Lines of sight are then chosen from random positions in the catalogue, and for each line of
sight we reconstruct the convergences from every halo within 5 arcminutes and compare the reconstructed
kappas with those from the ray tracing results. The standard deviation on  
shown in Figure \ref{fig:ScattervsTruncation}. It is clear from Figure \ref{fig:ScattervsTruncation} 
that as long as the truncation is greater than a few virial radii, the choice of truncation
radius does not broadly affect the scatter on $\kapparay$-$kapparec$; from now on we choose 
to truncate the halos at ten times their virial radius.
